Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2020-06-03 source/sec_intro.tex
intro updates

% Gaps in the theoretical understanding. 
To address these challenges, Federated Averaging (FedAvg)~\cite{mcmahan2016communication} was proposed as a particularly effective heuristic, which has enjoyed great empirical success~\cite{47586}. 
Its success has since motivated a growing line of research efforts into understanding its theoretical convergence guarantees in various settings.

 Existing convergence guarantees of FedAvg were either largely established on simplified settings~\cite{stich2018local,khaled2019first,wang2018cooperative,yu2019parallel} or provides a suboptimal 
convergence rate~\cite{li2019convergence}. Furthermore, the convergence of
FedAvg's accelerated variants~\cite{yu2019linear,huo2020faster,liu2019accelerating} under the realistic setting is yet to be analyzed.

% Our contributions
% \subsection{Contributions}
\textbf{Contributions.}
This work focuses on providing comprehensive convergence analysis
of FedAvg and its accelerated variants under a realistic setting.
Our results show
that the less number of communication rounds is needed to converge to a
target accuracy while preserving the linear speedup w.r.t. the number of participating devices.
More specifically, for strongly convex and smooth problems, we show 
that FedAvg admits a convergence rate of 
{\small{$\cO(1/NT)$}}, and local steps can be as large as {\small{$\cO(\sqrt{T/N})$}}. For convex and smooth objectives, we show the rate of {\small{$\cO(1/\sqrt{NT})$}} and local steps can be as large as {\small{$\cO(\sqrt{T/N})$}}. 
Moreover, we show that FedAvg with Nesterov acceleration also enjoys the same rates and speedup. 
%The same conclusions are equally applied for Nesterov accelerated FedAvg. 
To the best of our knowledge, this is the first linear speedup results for FedAvg with Nesterov accelerated SGD. Furthermore, 
we provide the first improved convergence rate for Nesterov accelerated FedAvg comparing to FedAvg under the popular overparamterized setting. 
Last but not least, we systematically examine the convergence rate of
FedAvg and its accelerated variants on both a synthetic dataset and  a real-world dataset. The empirical observations are well-aligned with the obtained theoretical findings.

@@

\textbf{Notations.}
Throughout the paper, we adopt the following notations. Let $N$ be the total
number of local devices, and $K \leq N$ is the maximal number of devices
that are accessible to the central server.  Let $T$ be the total number of SGD iterations performed by each local device, $E$ be the local steps between two consecutive server communications. Thus, $T/E$ is
the number of communications.

>>>>

% Gaps in the theoretical understanding. 
To address these challenges, Federated Averaging (FedAvg)~\cite{mcmahan2016communication} was proposed as a particularly effective heuristic, which has enjoyed great empirical success~\cite{47586}. Its success has since motivated a growing line of research efforts into understanding its theoretical convergence guarantees in various settings. Most existing convergence guarantees of FedAvg were either established in simplified settings~\cite{stich2018local,khaled2019first,wang2018cooperative,yu2019parallel,khaled2020tighter,li2018federated}, have restrictive assumptions~\cite{haddadpour2019convergence} or provide a suboptimal 
convergence rate~\cite{li2019convergence}. Furthermore, the convergence of
FedAvg's accelerated variants~\cite{yu2019linear,huo2020faster,liu2019accelerating} in the general heterogeneous and convex setting is yet to be analyzed. At the heart of the success of FL is the \emph{linear speedup} property with respect to the number of participating clients that can be achieved with infrequent communications. Although results of this kind are abundant in distributed optimization~\cite{yu2019linear,yu2019parallel,stich2018local,khaled2020tighter,stich2018local}, generalization to the FL setting with both types of heterogeneity is currently lacking.

\subsection{Related Works}
\textbf{The Convergence of FedAvg and Linear Speedup.}
There have been a considerable amount of convergence results in distributed stochastic optimization with infrequent communications that establish linear speedup in the number of clients. However, most of these prior works provide
convergence guarantees under simplified settings, where either statistical
heterogeneity~\cite{stich2018local,zhou2017convergence,wang2018cooperative,woodworth2018graph} or system
heterogeneity~\cite{yu2019parallel,wang2019adaptive,khaled2019first,jiang2018linear} are not
considered, and the associated algorithm is often referred to as local SGD. ~\cite{li2019convergence} are the first to provide convergence of FedAvg for strongly convex problems considering both statistical heterogeneity and system heterogeneity, yet it
didn't achieve linear speedup \cite{yu2019parallel,haddadpour2019convergence,stich2018local}. In~\cite{haddadpour2019convergence,liang2019variance,huo2020faster,jiang2018linear}, the convergence of FedAvg for non-convex problems has been studied, whereas this paper focuses on convex settings.
\begin{comment}
However, \cite{huo2020faster} did not discuss the linear speedup and the theorems in \cite{haddadpour2019convergence,liang2019variance} failed to cover the accelerated version of
FedAvg. 
\end{comment}

\textbf{Accelerating Stochastic Federated Learning Algorithms.}
% The accelerated SGD methods has achieved optimal convergence rate in strongly convex
% optimization~\cite{nesterov1983method}.
The convergence of accelerated FL algorithms using momentum (Nesterov and Heavy Ball) updates has been studied mostly in non-convex settings~\cite{yu2019linear,huo2020faster}. \cite{yu2019linear} does not consider system heterogeneity, while the result in \cite{huo2020faster} does not have linear speedup. Furthermore, the rates cannot improve over SGD-based algorithms in general. \cite{liu2019accelerating} provides an acceleration result for FedAvg with accelerated GD. \cite{liang2019variance} considers variance reduction based acceleration of FedAvg. This paper considers accelerated FedAvg with momentum-based stochastic updates in general and overparameterized convex settings.

\begin{comment}
	In this work, we provide a novel accelerated FedAvg with improved convergence rates under a popular overparameterized setting. 
\end{comment}

% Our contributions
% \subsection{Contributions}
\textbf{Contributions.}
We provide a comprehensive convergence analysis
of FedAvg and its accelerated variants in convex stochastic problems with both system and statistical heterogeneity.
Our results demonstrate the desirable linear speedup property with respect to the number of participating clients.
More specifically, for strongly convex and smooth problems, we show 
that FedAvg admits a convergence rate of 
{\small{$\cO(1/NT)$}}, and local steps can be as large as {\small{$\cO(\sqrt{T/N})$}}. For convex and smooth objectives, the rate is {\small{$\cO(1/\sqrt{NT})$}}. 
Moreover, we show that FedAvg with Nesterov acceleration also enjoys the same rates and speedup. 
%The same conclusions are equally applied for Nesterov accelerated FedAvg. 
To the best of our knowledge, this is the first linear speedup result for FedAvg and accelerated variants in the full heterogeneous FL setting. Furthermore, 
we propose a new variant of FedAvg and provide the first improved convergence rate for stochastic momentum-accelerated FedAvg in the popular overparamterized setting.
Last but not least, we systematically examine the convergence rate of
FedAvg and its accelerated variants on both a synthetic dataset and  a real-world dataset. The empirical observations are well-aligned with the obtained theoretical findings.

@@

\textbf{Notations.}
Throughout the paper, we adopt the following notations. Let $N$ be the total
number of local devices, and $K \leq N$ is the maximal number of devices
that are accessible to the central server.  Let $T$ be the total number of stochastic updates performed by each local device, $E$ be the local steps between two consecutive server communications. Thus, $T/E$ is
the number of communications.
